Measuring the Type Ia supernova delay time distribution via host galaxy rates in the Dark Energy Survey

### 2021-01-12 draft.tex
Updates from Overleaf

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Rates of SNe Ia in DES]{Supernova Host Galaxies in the Dark Energy Survey: III. Constraining the SN Ia delay time dstribution}

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_all.png}
    \caption{The rate per galaxy of SNe Ia as a function of stellar mass.}%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_raw}
\end{figure}
%%%% INCOMPLETENESS %%%%
\section{Incompleteness corrections}
\label{sec:incompleteness}
Although Eq. \ref{eq:rate1} serves as an approximation to the SN rate per galaxy per year, both the SN and field galaxy samples introduced in Sections \ref{subsec:host_sample} and \ref{subsec:field_sample} are affected by incompleteness, which is likely to be the dominant systematic effect in the analysis. In this section we describe our method of correcting for various sources incompleteness in the data.

@@

%%% MODEL %%%
\section{Modelling the rate of type Ia supernovae}
 \label{sec:method}
We represent the SN Ia DTD by a power law of the form:
\begin{equation}
    \Phi(\tau) = A\left(\frac{\tau}{\mathrm{Gyr}}\right)^{\beta}\,,
\end{equation} 
where $\tau$ is the time since a burst of star formation, and $A$ is a normalisation at $\tau=1\mathrm{Gyr}$. If all stars in a galaxy were formed at a single epoch $t_f$, the DTD would describe the rate of SNe at an observation time $\tau = t_0$. However in reality galaxies are formed by the gradual build up of stellar mass over several epochs of star formation -- the distribution of this mass build up is known as the star-formation history (SFH). The rate of SNe Ia is thus the sum of the DTD evaluated for each epoch of star formation, multiplied by the stellar mass formed in that epoch. Mathematically this is represented by the convolution of the DTD and SFH:

@@

@@

The rate of SNe Ia per galaxy as a function of stellar mass is shown in Fig. \ref{fig:rate_raw}. The relationship between SN Ia rate and galaxy stellar mass is well described by a linear function. To find the slope and intercept that best describe the data we employ Hamiltonian Monte-Carlo (HMC), which we implement via the \texttt{pystan} package and describe in full detail in Appendix \ref{appendix:linear_fits}. 
\subsection{Dependence of the SN Ia rate on SFR \label{subsec:rates_sfr}}

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Rates of SNe Ia in DES]{Measuring the Type Ia supernova delay time distribution via host galaxy rates in the Dark Energy Survey}

@@

%%%% INCOMPLETENESS %%%%
\section{Incompleteness corrections}
\label{sec:incompleteness}
Although Eq. \ref{eq:rate1} serves as an approximation to the SN rate per galaxy per year, both the SN and field galaxy samples introduced in Sections \ref{subsec:host_sample} and \ref{subsec:field_sample} are affected by incompleteness, which is likely to be the dominant systematic effect in the analysis. In this section we describe our method of correcting for various sources incompleteness in the data.

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_all.png}
    \caption{The rate per galaxy of SNe Ia as a function of stellar mass.}%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_raw}
\end{figure}

%\begin{figure}
%    \centering
%    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_split_sfr.png}
%    \caption{The rate per galaxy of SNe Ia as a function of stellar mass for moderately and highly star-forming galaxies, as well as passive galaxies.}%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
%    \label{fig:rate_raw_split_sfr}
%\end{figure}
%%% MODEL %%%
\section{Modelling the rate of type Ia supernovae}
 \label{sec:method}
We represent the SN Ia DTD by a power law. We consider the DTD effective after some "prompt time" $t_{\mathrm{p}}$, before which its value is set to 0. This is generally interpreted as the time taken for WDs to form after the burst of star formation. The DTD is thus:
\begin{equation}
 \Phi(\tau) = \left\{
    \begin{array}{@{}cc@{}}
    0 , & \tau < t_{\mathrm{p}} \\
    A\left(\frac{\tau}{\mathrm{Gyr}}\right)^{\beta} , & \tau \geq t_{\mathrm{p}}
    \end{array}\right.
        \label{eq:dtd}
\end{equation} 
where $\tau$ is the time since a burst of star formation, and $A$ is a normalisation at $\tau=1$ Gyr. If all stars in a galaxy were formed at a single epoch $t_f$, the DTD would describe the rate of SNe at an observation time $\tau = t_0$. However in reality galaxies are formed by the gradual build up of stellar mass over several epochs of star formation -- the distribution of this mass build up is known as the star-formation history (SFH). The rate of SNe Ia is thus the sum of the DTD evaluated for each epoch of star formation, multiplied by the stellar mass formed in that epoch. Mathematically this is represented by the convolution of the DTD and SFH:

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_all.png}
    \caption{The rate per galaxy of SNe Ia as a function of stellar mass.}%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_raw}
\end{figure}

@@

The rate of SNe Ia per galaxy as a function of stellar mass is shown in Fig. \ref{fig:rate_raw}. The relationship between SN Ia rate and galaxy stellar mass is well described by a linear function in log space, which corresponds to a power law. To find the slope and intercept that best describe the data we employ Hamiltonian Monte-Carlo (HMC), which we implement via the \texttt{pystan} package and describe in full detail in Appendix \ref{appendix:linear_fits}. We measure a slope of $0.70\pm0.02$, which is broadly consistent with the values found in SNLS and SDSS by \citet{Sullivan2006} and \citet{Smith2012} respectively. It is evident that the linear fit is not consistent with the data across the full range of stellar mass considered. In particular, there appear to be two locations of interest: firstly, the slope is steeper at masses lower than $10^9$ M$_{\odot}$; secondly there is an apparent flattening and possibly a turnover at masses around and above $10^11$ M$_{\odot}$. The most likely cause of these divergences is because the SN rate is driven by the SFH which is not a linear function of stellar mass (see Section \ref{sec:method}). Second order effects may also be important, and we address these in Section \ref{sec:discussion}.

%\textbf{TEMPTED TO DITCH THIS SINCE OUR SFR IS SKETCHY}
%In order to make a more detailed comparison of our results with those of \citet{Sullivan2006} and \citet{Smith2012}, we split galaxies according to their sSFR: galaxies with sSFR$\leq -11.5$ yr$^{-1}$ are denoted as passive, those with $-11.5 < \mathrm{sSFR} \leq -9.5$ are moderately star-forming, and those with sSFR $>-9.5$ are labelled as highly star-forming. The results when split this way are shown in Fig. \ref{fig:rate_raw_split_sfr}. Star-forming galaxies show a steeper slope than their passive counterparts, with values of $1.10 \pm 0.05$ and $0.84 \pm 0.04$ for moderate and highly star-forming galaxies respectively, while passive galaxies show a much shallower slope of $0.55 \pm 0.08$. These values are broadly similar to those measured in SDSS by \citet{Smith2012}, with star-forming galaxies showing steeper slopes than passives (\citet{Smith2012} found 1.01 and 0.72 for SF and passive, respectively). We note that we detect no passive galaxies with mass below $10^{10} \mathrm{M}_{\odot}$. Above this value, the \citet{Smith2012} data for passive galaxies is considerably flatter than that which we measure. On the other hand our results are inconsistent with those found at a similar redshift to the DES data in SLNS \citep{Sullivan2006}, who measured a slope of 1.10 in passives but 0.66 and 0.74 in moderate and highly star-forming galaxies respectively. We note that the moderately star-forming slope is thus consistent between DES and SNLS, but the slopes of passive and highly star-forming galaxies are swapped.

\subsection{Constraints on the SN Ia delay time distribution}
\label{subsec:results_dtd}

In Section \ref{sec:method} we showed how the rate of SNe Ia is driven by the convolution of the SFH of galaxies and the SN Ia DTD (Eq. \ref{eq:galaxy_rate}). We then prescribed a model to infer the mean SFHs of galaxies of any given stellar mass. In this section, we measure the DTD by forward modelling it through Eq. \ref{eq:galaxy_rate} at the stellar masses corresponding to the centres of the bins in Fig. \ref{fig:rate_raw}.

We assume a DTD $\Phi(\tau)$ with a power-law form described by index $\beta$ (Eq. \ref{eq:dtd}), normalisation $A$ and effective after prompt time $t_{\mathrm{p}}$. As per Section \ref{subsec:rates_mass}, we constrain the parameters via Bayesian inference, using HMC. At each step of the sampling procedure the model is re-calculated via Eq. \ref{eq:galaxy_rate}, and evaluate the likelihood assuming that the rate data are normally distributed about their means. For $\beta$ we adopt a Gaussian prior with hyper-parameters $p(\beta) \sim \mathcal{N}(-1\,,0.3)$. We fit for $A$ in log space, and adopt a Gaussian prior (in log space) with hyper-parameters $p(A) \sim \mathcal{N}(-12.7\,, 0.5)$.

SNe Ia do not occur instantaneously after an episode of star formation. Instead stars with zero-age main-sequence (ZAMS) masses less than $8 \mathrm{M}_{\odot}$ take time to evolve along the main sequence and form WDs and this time is not well constrained. In many works this ``prompt time" $t_{\mathrm{p}}$ is fixed to some value expected to be the minimum time for a star to evolve off the MS and become a WD, such as 40 Myr \citep{Maoz2012,Graur2014,Graur2018}. Other studies have left $t_{\mathrm{p}}$ as a free parameter \citep{Heringer2019,Asier2020}. We perform two fits: one with $t_{\mathrm{p}}$ fixed at 40 Myr, and one with it as a free parameter. We fit $t_{\mathrm{p}}$ in log space and adopt a Gaussian prior with hyper-parameters $p(\log(t_{\mathrm{p}})) \sim \mathcal{N}(-1.3\,,0.2)$, which corresponds to being centred around 50 Myr. Choosing this regime to fit allows more prior weight to be placed on shorter prompt times as per the majority of the literature.

The posterior distributions for the fit parameters are found in Figs. \ref{fig:corner_beta_norm} and \ref{fig:corner_beta_norm_tp} for the fixed and free $t_{\mathrm{p}}$ fits respectively, and the posterior means and standard deviations are summarised in Table \ref{tab:dtd_results}. The inclusion of $t_{\mathrm{p}}$ in the fit produces results for all three parameters that are significantly different than in the fixed case, suggesting either that the prompt time is different to 40 Myr, or that the method is incapable of constraining these three parameters simultaneously.
